Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-30 text/manuscript-full-KS.qmd
results revisions

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households. The model created a classification system that grouped SAT test-takers into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. @RN4982[p. 4] comment that the Market Segment Model "was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways."

@@

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail [CRYSTAL]. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.^[This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."] Based on these analyses, @RN4982[p. 46] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->
```

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.^[@RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

@@

Quantification "disciplines" [@RN5007] actors to react in particular ways. EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges. Anecdotally, the salience of Geomarkets and EPS software affected the decision-making authority of admissions officers and the territories that admissions officers were assigned to.^[Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago [ADD TO APPENDIX], of Lake Forest College, of Saint Mary's College [ADD TO APPENDIX CRYSTAL]] <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

@@

```{=html}
<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->
```

We also expect substantial racial inequality between Geomarkets within metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized by extreme historic and contemporary residential racial segregation [@RN4779]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the contours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. <!--#  Structures built upon racialized structures are racialized structures [RN4786].-->

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.^[For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to names.] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability.^[Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and low-income students.] <!--^[Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%).] --> @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years.^[20.6% of students who participated in Search obtained a BA in four years compared to 15.7% of students who opted out, representing a 31.2% (=(20.6-15.7)/15.7) increase in the relative probability of graduation.] <!-- , particularly for sub-groups that were historically excluded from college.^[. Furthermore, the relative increase in the probability of obtaining a BA was higher for Black (40.3%), Hispanic (43.3%), and Native American/Alaska Native students (48.3%) than it was for White (27.9%) and Asian (28.2%) students. The relative increase was also higher for students whose parents did not attend college (50.0%) than it was for students whose parents had a BA (25.3%).]-->

@@

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, such as Geomarkets and Geodemographic Segment filters.^[Geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565].] In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." <!--# A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score.--> We extend @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets. <!--# This manuscript focuses on Geomarkets. -->

@@

We use data from the 1980 and the 2000 Decennial Census as well as 5-year estimates from the 2020 American Community Survey (ACS). For the 1980 Decennial Census, we draw from Summary Tape File 1 (STF1) for variables about race and Hispanic origin collected via "short form" questions answered by all households. We draw from Summary Tape File 3 (STF3) for variables about socioeconomic characteristics (income, education, and poverty) collected via the "long form" questionnaire completed by a sample of roughly 20% of all households. The 1980 Decennial Census data collection period closely matches the data collection period of SAT score-sending behavior that @RN4982 utilized to create Geomarkets and the Market Segment Model. For the 2000 Decennial Census, we similarly use the Summary File 1A (SF1A) for census-tract measures of race and Hispanic origin and the Summary File 3A (SF3A) for census-tract measures of income, education, and poverty.^[Data for the 1980 and 2000 Decennial Census were retrieved from the IPUMS National Historical Geographic Information System, created by @RN5028.] Beginning in 2010, the Decennial Census no longer collected a "long form" questionnaire. These questions were replaced by ACS. Therefore, we also utilized the 2020 5-year ACS for both race and socioeconomic characteristics at the census tract level, which includes data collected from 2016-2020.^[Data for the 2020 5-year ACS were retreived from the [tidycensus](https://walker-data.com/tidycensus/) R package, created by @tidycensus_citation.] This data collection period mirrors the period for our primary data collection about student list purchases used to analyze RQ2, as described below.

@@

We assigned each census tract to College Board Geomarkets by implementing a partial spatial merge. Spatial manipulations utilized functions from the *sf* R package. Data to create the shapefiles for the College Board Geomarkets were obtained from a 2012 R-bloggers post written by a data scientist working in the enrollment management space.^[The author of the post reported that he found the zip codes associated with Geomarkets from a Google search.] Essentially, the data file had one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the `aggregate()` function to create shapefiles for each Geomarket.

@@

The primary data source for our second research question is student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Texas, and Arizona. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase. The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters.^[The caveat to this statement is that most purchases select to exclude prospect profiles that were previously purchased by the university.] Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter and purchased prospects from all Geomarkets in a particular metropolitan area.^[For analyses of Dallas, Detroit, Long Island, Northern New Jersey, and Philadelphia, we utilize data from student list purchases that did filter on Geomarkets but included all Geomarkets in the region.] For each prospect, we know their home zip code and, therefore, which Geomarket they belong to. Our analyses then show which students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

The variables of interest are Geomarket and prospect-level characteristics, which are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity non-response) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016) \[CITE\]. From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016) \[CITE\].^[Any student who selects a Hispanic ethnicity category is defined as Hispanic/Latino, regardless of the race categories they select. Additionally, non-Hispanic students who check “American Indian or Alaska Native” and another race group are defined as "two or more races, non-Hispanic."] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].

>>>>

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households. The model created a classification system that grouped SAT test-takers into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. @RN4982 [p. 4] comment that the Market Segment Model "was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways."

@@

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```
The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail \[CRYSTAL\]. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

[^5]: For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.

[^6]: Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.[^7] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

[^7]: This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."

@@

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->
```
Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^8] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

[^8]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

@@

Quantification "disciplines" [@RN5007] actors to react in particular ways. EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.[^9] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges. Anecdotally, the salience of Geomarkets and EPS software affected the decision-making authority of admissions officers and the territories that admissions officers were assigned to.[^10] <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

[^9]: Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."

[^10]: Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago \[ADD TO APPENDIX\], of Lake Forest College, of Saint Mary's College \[ADD TO APPENDIX CRYSTAL\]

@@

```{=html}
<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->
```
We also expect substantial racial inequality between Geomarkets within metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized by extreme historic and contemporary residential racial segregation [@RN4779]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the contours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. <!--#  Structures built upon racialized structures are racialized structures [RN4786].-->

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

[^11]: For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to names.

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability.[^12] <!--^[Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%).] --> @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years.[^13] <!-- , particularly for sub-groups that were historically excluded from college.^[. Furthermore, the relative increase in the probability of obtaining a BA was higher for Black (40.3%), Hispanic (43.3%), and Native American/Alaska Native students (48.3%) than it was for White (27.9%) and Asian (28.2%) students. The relative increase was also higher for students whose parents did not attend college (50.0%) than it was for students whose parents had a BA (25.3%).]-->

[^12]: Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and low-income students.

[^13]: 20.6% of students who participated in Search obtained a BA in four years compared to 15.7% of students who opted out, representing a 31.2% (=(20.6-15.7)/15.7) increase in the relative probability of graduation.

@@

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, such as Geomarkets and Geodemographic Segment filters.[^14] In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." <!--# A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score.--> We extend @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets. <!--# This manuscript focuses on Geomarkets. -->

[^14]: Geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565].

@@

We use data from the 1980 and the 2000 Decennial Census as well as 5-year estimates from the 2020 American Community Survey (ACS). For the 1980 Decennial Census, we draw from Summary Tape File 1 (STF1) for variables about race and Hispanic origin collected via "short form" questions answered by all households. We draw from Summary Tape File 3 (STF3) for variables about socioeconomic characteristics (income, education, and poverty) collected via the "long form" questionnaire completed by a sample of roughly 20% of all households. The 1980 Decennial Census data collection period closely matches the data collection period of SAT score-sending behavior that @RN4982 utilized to create Geomarkets and the Market Segment Model. For the 2000 Decennial Census, we similarly use the Summary File 1A (SF1A) for census-tract measures of race and Hispanic origin and the Summary File 3A (SF3A) for census-tract measures of income, education, and poverty.[^15] Beginning in 2010, the Decennial Census no longer collected a "long form" questionnaire. These questions were replaced by ACS. Therefore, we also utilized the 2020 5-year ACS for both race and socioeconomic characteristics at the census tract level, which includes data collected from 2016-2020.[^16] This data collection period mirrors the period for our primary data collection about student list purchases used to analyze RQ2, as described below.

[^15]: Data for the 1980 and 2000 Decennial Census were retrieved from the IPUMS National Historical Geographic Information System, created by @RN5028.

[^16]: Data for the 2020 5-year ACS were retreived from the [tidycensus](https://walker-data.com/tidycensus/) R package, created by @tidycensus_citation.

@@

We assigned each census tract to College Board Geomarkets by implementing a partial spatial merge. Spatial manipulations utilized functions from the *sf* R package. Data to create the shapefiles for the College Board Geomarkets were obtained from a 2012 R-bloggers post written by a data scientist working in the enrollment management space.[^17] Essentially, the data file had one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the `aggregate()` function to create shapefiles for each Geomarket.

[^17]: The author of the post reported that he found the zip codes associated with Geomarkets from a Google search.

@@

The primary data source for our second research question is student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Texas, and Arizona. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase. The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters.[^18] Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^18]: The caveat to this statement is that most purchases select to exclude prospect profiles that were previously purchased by the university.

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter and purchased prospects from all Geomarkets in a particular metropolitan area.[^19] For each prospect, we know their home zip code and, therefore, which Geomarket they belong to. Our analyses then show which students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

[^19]: For analyses of Dallas, Detroit, Long Island, Northern New Jersey, and Philadelphia, we utilize data from student list purchases that did filter on Geomarkets but included all Geomarkets in the region.

The variables of interest are Geomarket and prospect-level characteristics, which are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity non-response) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016) \[CITE\]. From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016) \[CITE\].[^20] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].
